Classical Dynamics of Harmonically Trapped Interacting Particles

### 2018-01-27 draft_11.tex
minor revisions, references added

The presence of harmonic trapping is thus a paradigmatic aspect in the field of quantum
non-equilibrium physics.  This motivates our study of classical dynamics of trapped interacting
particles.  Similarly motivated by quantum dynamics, classical trapped dynamics of hard rods has
recently been studied in Refs.\ \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10}, focusing on
integrability-breaking effects of the trap.  In this work, we consider trapped classical particles
which interact via a simple finite-range potential between any pair of particles: a constant
repulsive force acts whenever the particles are less than a certain distance apart.  The interaction
strength can be tuned, leading to different regimes of behavior.

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interaction, the breathing
mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.  A mean
field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the trapping
frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the mean-field
description is expected to be qualitatively correct, thus the breathing mode frequency is expected
to change monotonically from $2\omega_0$ at zero interactions to $\sqrt{5}\omega_0$ at large
interactions.  Remarkably, for 2D traps, the breathing frequency is exactly $2\omega_0$ for any
interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical trapped particles
with scale-invariant interaction potentials also have breathing frequency $2\omega_0$ independent of
interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped
Lieb-Liniger gas) is also remarkable: as the interaction is increased, the breathing frequency first
decreases from $2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger
interactions increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau)
limit \cite{Naegerl_Science09, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{OrignacCitro_PRA08, br_mode_long_range_interactions, Bonitz_Review}.  The
breathing mode has also been studied for a high-temperature trapped gas where a Boltzmann-equation
description is valid, both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

@@

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represent the thermal value
\cite{PolkovnikovRigol_AdvPhys2016, BorgonoviIzrailevSantos_PhysRep2016,
  ETH_Deutsch_Srednicki_Rigol}.  The connection to statistical mechanics is particularly difficult
for explicitly finite systems, in which case ETH is to be understood in terms of finite size scaling
\cite{Beugeling_ETHscaling_PRE14}.  In classical systems, the connection between microscopic
dynamics and statistical mechanics has been continually studied since the 19th century
\cite{Gaspard_book_1998, Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book,
  EckmannRuelle_RMP85, Gaspard_PhysicaA06}.  It is generally understood that systems with nonzero
Lyapunov exponents, i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided
there are enough degrees of freedom.  As in the quantum case, finite systems are particularly
intricate.  In finite systems, defining thermalization is trickier as most observables will
fluctuate or oscillate substantially.  In addition, the influence of Kolmogorov-Arnold-Moser (KAM)
orbits may play a role in preventing thermalization; it is generally believed that the fraction of
phase space where KAM physics might be relevant decreases rapidly with particle number
\cite{Dumas_book_KAMstory}.

In our work, we examine the relaxation of our trapped interacting classical system, explicity for
finite numbers of particles.  We look for thermalization within finite timescales, by sampling
single-particle energies periodically until such timescales and comparing the distribution of
single-particle energies with the Boltzmann distribution.  We show that whether or not the system
thermalizes in this sense depends on the ratio of interaction strength and energy per particle.  We
also show that this result is reflected in the finite-time Lyapunov exponents of the system.
%
We also examine another notion of thermalization, according to which a system is thermalized when it
loses memory of initial state, by following the dynamics of the particle distribution in phase
space.  Remarkably, we find that one can observe energy thermaliztion even at timescales when the
shape of the phase space distribution continues to perform seemingly coherent oscillations.

@@

Our model for the interacting classical gas involves particles with a simple finite-range repulsive
interaction.  Two particles repel each other whenever they are within a distance $\sigma$ from each
other.  The force of repulsion is a constant, $F_0$, within this distance and zero when the distance
is larger. 
%
\begin{equation}
\Big| F(x) \Big| ~=~ \begin{cases}
F_0  &\quad |x|<\sigma\\
0  &\quad |x|>\sigma
\end{cases}
\label{eq:preparation1}
\end{equation}
where $|x|$ is the distance between two particles.  This equation describes the magnitude of the
interparticle force; the direction is always repulsive.

@@

\bibitem{Guery-Odelin1999}
D.~Gu\'ery-Odelin, F.~Zambelli, J.~Dalibard, and S.~Stringari, ``Collective
  oscillations of a classical gas confined in harmonic traps,'' {\em Phys. Rev.
  A}, {\bf 60}, 4851 (1999).

@@

\bibitem{Beugeling_ETHscaling_PRE14}
W. Beugeling, R. Moessner, and M.~Haque, 
``Finite-size scaling of eigenstate thermalization'', 
{\em Phys. Rev. E}, {\bf 89}, 042112 (2014).

@@

\bibitem{deWijn_fine}
A.~S.~de~Wijn, B.~Hess, and B.~V.~Fine, 
``Largest Lyapunov Exponents for Lattices of Interacting Classical Spins'', 
{\em Phys. Rev. Lett.} {\bf 109}, 034101 (2012). 
%
\\
%
A.~S.~de~Wijn, B.~Hess, and B.~V.~Fine, 
``Chaotic properties of spin lattices near second-order phase transitions'',
{\em Phys. Rev. E}, {\bf 92}, 062929 (2015). 
%
\\
%
A.~S.~de~Wijn, B.~Hess, and B.~V.~Fine, 
``Lyapunov instabilities in lattices of interacting classical spins at infinite temperature'', 
{\em J.~Phys.~A: Math. Theor.}, {\bf 46}, 254012 (2013).

\bibitem{Lyapunov_numerical_calculations}
C.~Dellago and H.~A.~Posch, 
``Kolmogorov-Sinai entropy and Lyapunov spectra of a hard-sphere gas''  
{\em Physica A},  {\bf 240}, 68 (1997).
%
\\
%
V,~Latora, A.~Rapisarda, and S.~Ruffo, 
``Lyapunov Instability and Finite Size Effects in a System with Long-Range Forces'', 
{\em Phys. Rev. Lett.}, {\bf 80}, 692 (1998). 
%
\\
%
W.~G.~Hoover, H.~A.~Posch, C.~Forster, C.~Dellago, and M.~Zhou, 
``Lyapunov Modes of Two-Dimensional Many-Body Systems; Soft Disks, Hard Disks, and Rotors'', 
{\em J. Stat. Phys.} , {\bf 109}, 765 (2002).

@@

\bibitem{FiniteTimeLyapunov} 
E.~Ott, ``Chaos in Dynamical Systems'', Cambridge University Press,  1993.
%
\\ 
%
M.~A.~Sep\'ulveda, R~Badii, and E.~Pollak, 
``Spectral analysis of conservative dynamical systems'', 
{\em Phys. Rev. Lett.}, {\bf 63}, 1226 (1989). 
%
\\ 
%
C.~Amitrano and R.~S.~Berry. 
``Probability distributions of local Lyapunov exponents for Hamiltonian systems'', 
{\em Phys. Rev. E}, {\bf 47}, 3158 (1993). 
%
\\ 
%
F.~K.~Diakonos, D.~Pingel, and P.~Schmelcher
``Analyzing Lyapunov spectra of chaotic dynamical systems'', 
{\em Phys. Rev. E}, {\bf 62}, 4413 (2000).
%
\\ 
%
H.~Schomerus and M.~Titov, 
``Statistics of finite-time Lyapunov exponents in a random time-dependent potential'', 
{\em Phys. Rev. E}, {\bf 66}, 066207 (2002). 
%
\\ 
%
M.~W.~Beims, C.~Manchein, and J.~M.~Rost, 
``Origin of chaos in soft interactions and signatures of nonergodicity'', 
{\em Phys. Rev. E}, {\bf 76}, 056203 (2007).
%
\\ 
%
C.~Manchein, M.~W.~Beims, and J.~M.~Rost, ``Characterizing the dynamics of higher dimensional
nonintegrable conservative systems'', {\em Chaos}, {\bf 22}, 033137 (2012).
%
\\ 
%
S.~Sawada and T.~Taniguchi, 
``Chaos and ergodicity of two hard disks within a circular billiard'', 
{\em Phys. Rev. E}, {\bf 88}, 022907 (2013). 
%
\\ 
%
K.~Kanno and A.~Uchida, ``Finite-time Lyapunov exponents in time-delayed nonlinear dynamical
systems'',  {\em Phys. Rev. E}, {\bf 89}, 032918 (2014). 
%
\\ 
%
D,~Paz\'o, J.~M.~L\'opez, and A.~Politi, 
``Diverging Fluctuations of the Lyapunov Exponents'' 
{\em Phys. Rev. Lett.},  {\bf 117}, 034101 (2016).

>>>>

The presence of harmonic trapping is thus a paradigmatic aspect in the field of quantum
non-equilibrium physics.  This motivates our study of classical dynamics of trapped interacting
particles.  Similarly motivated by quantum dynamics, classical trapped dynamics of hard rods has
recently been studied in Refs.\ \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10}, focusing on
integrability-breaking effects of the trap.  Classical dynamics in parabolic traps have also been
studied for particles with long-range interactions, for which shell structures play an important
role \cite{Peeters_shell}.  
%
In this work, we consider trapped classical particles which interact via a simple finite-range
potential between any pair of particles: a constant repulsive force acts whenever the particles are
less than a certain distance apart.  The interaction strength can be tuned, leading to different
regimes of behavior.

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interaction, the breathing
mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.  A mean
field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the trapping
frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the mean-field
description is expected to be qualitatively correct, thus the breathing mode frequency is expected
to change monotonically from $2\omega_0$ at zero interactions to $\sqrt{5}\omega_0$ at large
interactions.  Remarkably, for 2D traps, the breathing frequency is exactly $2\omega_0$ for any
interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical trapped particles
with scale-invariant interaction potentials also have breathing frequency $2\omega_0$ independent of
interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped
Lieb-Liniger gas) is also remarkable: as the interaction is increased, the breathing frequency first
decreases from $2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger
interactions increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau)
limit \cite{Naegerl_Science09, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{OrignacCitro_PRA08, br_mode_long_range_interactions, Bonitz_Review}.  The
breathing mode has also been studied for a high-temperature trapped gas where a Boltzmann-equation
description is valid, both theoretically \cite{GueryOdelin1999, GueryOdelin_thermalization_PRA06}
and experimentally \cite{Cornell_classsical_3Dexpt_NatPhys15}.

@@

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represent the thermal value
\cite{PolkovnikovRigol_AdvPhys2016, BorgonoviIzrailevSantos_PhysRep2016,
  ETH_Deutsch_Srednicki_Rigol}.  The connection to statistical mechanics is particularly difficult
for explicitly finite systems, in which case ETH is to be understood in terms of finite size scaling
\cite{ETHscaling}.  In classical systems, the connection between microscopic dynamics and
statistical mechanics has been studied since the 19th century \cite{Gaspard_book_1998,
  Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85,
  Gaspard_PhysicaA06}.  It is generally understood that systems with nonzero Lyapunov exponents,
i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided there are enough
degrees of freedom.  
%
However, various aspects deserve further study.  For example, given a certain strength of
integrability-breaking (e.g., interaction strength), one might want to ask whether a system achieves
a thermal distribution under purely Newtonian (non-dissipative) evolution, and if so, what the
necessary timescale is for this thermalization.
%
As in the quantum case, finite systems are particularly intricate.  In finite systems, defining
thermalization is trickier as most observables will fluctuate or oscillate substantially.  In
addition, the influence of Kolmogorov-Arnold-Moser (KAM) orbits may play a role in preventing
thermalization; it is generally believed that the fraction of phase space where KAM physics might be
relevant decreases rapidly with particle number \cite{Dumas_book_KAMstory}.

In our work, we examine the relaxation of our trapped interacting classical system, explicity for
finite numbers of particles.  We look for thermalization within finite timescales, by sampling
single-particle energies periodically until such timescales and comparing the distribution of
single-particle energies with the Boltzmann distribution.  We show that whether or not the system
thermalizes in this sense depends on the ratio of interaction strength and energy per particle.  We
also show that this result is reflected in the Lyapunov exponents of the system.  (Since we include
weak ergodicity breaking for which thermalization takes ultra-long times, we find it convenient to
consider \emph{finite-time} Lyapunov exponents.)
%
We also examine another notion of thermalization, according to which a system is thermalized when it
loses memory of initial state, by following the dynamics of the particle distribution in phase
space.  Remarkably, we find that one can observe energy thermaliztion even at timescales when the
shape of the phase space distribution continues to perform seemingly coherent oscillations.

@@

Our model for the interacting classical gas involves particles with a finite-range repulsive
interaction.  Two particles repel each other whenever they are within a distance $\sigma$ from each
other.  The force of repulsion is a constant, $F_0$, within this distance and zero when the distance
is larger.
%
\begin{equation}
\Big| F(x) \Big| ~=~ \begin{cases}
F_0  &\quad |x|<\sigma\\
0  &\quad |x|>\sigma
\end{cases}
\label{eq:preparation1}
\end{equation}
where $|x|$ is the distance between two particles.  This equation describes the magnitude of the
interparticle force; the direction is always repulsive.

@@

\bibitem{Peeters_shell} V.~M.~Bedanov and and F.~M.~Peeters, ``Ordering and phase transitions of
  charged particles in a classical finite two-dimensional system'', {\em Phys. Rev. B}, {\bf 49},
  2667 (1994).
%
\\
%
V.~A.~Schweigert and F.~M.~Peeters, ``Time-dependent properties of classical
  artificial atoms'', {\em J. Phys.: Condens. Matter}, {\bf 10}, 2417 (1998).

\bibitem{GueryOdelin1999}
D.~Gu\'ery-Odelin, F.~Zambelli, J.~Dalibard, and S.~Stringari, ``Collective
  oscillations of a classical gas confined in harmonic traps,'' {\em Phys. Rev.
  A}, {\bf 60}, 4851 (1999).

@@

\bibitem{ETHscaling}
W. Beugeling, R. Moessner, and M.~Haque, 
``Finite-size scaling of eigenstate thermalization'', 
{\em Phys. Rev. E}, {\bf 89}, 042112 (2014). 
%
M.~Haque and P.~McClarty, ``Eigenstate Thermalization Scaling in Majorana Clusters: from Chaotic to
Integrable Sachdev-Ye-Kitaev Models'', arXiv:1711.02360.

@@

\bibitem{deWijn_fine}
A.~S.~de~Wijn, B.~Hess, and B.~V.~Fine, 
``Largest Lyapunov Exponents for Lattices of Interacting Classical Spins'', 
{\em Phys. Rev. Lett.} {\bf 109}, 034101 (2012). 
%
\\
%
A.~S.~de~Wijn, B.~Hess, and B.~V.~Fine, 
``Lyapunov instabilities in lattices of interacting classical spins at infinite temperature'', 
{\em J.~Phys.~A: Math. Theor.}, {\bf 46}, 254012 (2013). 
\\
%
T.~A.~Elsayed, B.~Hess, and B.~V.~Fine, 
``Signatures of chaos in time series generated by many-spin systems at high temperatures'',
{\em Phys. Rev. E}, {\bf 90}, 022910 (2014). 
%
\\
%
A.~S.~de~Wijn, B.~Hess, and B.~V.~Fine, 
``Chaotic properties of spin lattices near second-order phase transitions'',
{\em Phys. Rev. E}, {\bf 92}, 062929 (2015). 
%

\bibitem{Lyapunov_numerical_calculations}
C.~Dellago and H.~A.~Posch, 
``Kolmogorov-Sinai entropy and Lyapunov spectra of a hard-sphere gas''  
{\em Physica A},  {\bf 240}, 68 (1997).
%
\\
%
V,~Latora, A.~Rapisarda, and S.~Ruffo, 
``Lyapunov Instability and Finite Size Effects in a System with Long-Range Forces'', 
{\em Phys. Rev. Lett.}, {\bf 80}, 692 (1998). 
%
\\
%
W.~G.~Hoover, H.~A.~Posch, C.~Forster, C.~Dellago, and M.~Zhou, 
``Lyapunov Modes of Two-Dimensional Many-Body Systems; Soft Disks, Hard Disks, and Rotors'', 
{\em J. Stat. Phys.} , {\bf 109}, 765 (2002).
%
\\
%
A.~Das, S.~Chakrabarty, A.~Dhar, A.~Kundu, R.~Moessner, S.~S.~Ray, and S.~Bhattacharjee,
``Light-cone spreading of perturbations and the butterfly effect in a classical spin chain'',
arXiv:1711.07505.

@@

\bibitem{FiniteTimeLyapunov} 
E.~Ott, ``Chaos in Dynamical Systems'', Cambridge University Press,  1993.
%
\\ 
%
M.~A.~Sep\'ulveda, R~Badii, and E.~Pollak, 
``Spectral analysis of conservative dynamical systems'', 
{\em Phys. Rev. Lett.}, {\bf 63}, 1226 (1989). 
%
\\ 
%
C.~Amitrano and R.~S.~Berry. 
``Probability distributions of local Lyapunov exponents for Hamiltonian systems'', 
{\em Phys. Rev. E}, {\bf 47}, 3158 (1993). 
%
\\ 
%
F.~K.~Diakonos, D.~Pingel, and P.~Schmelcher
``Analyzing Lyapunov spectra of chaotic dynamical systems'', 
{\em Phys. Rev. E}, {\bf 62}, 4413 (2000).
%
\\ 
%
H.~Schomerus and M.~Titov, 
``Statistics of finite-time Lyapunov exponents in a random time-dependent potential'', 
{\em Phys. Rev. E}, {\bf 66}, 066207 (2002). 
%
\\ 
%
M.~W.~Beims, C.~Manchein, and J.~M.~Rost, 
``Origin of chaos in soft interactions and signatures of nonergodicity'', 
{\em Phys. Rev. E}, {\bf 76}, 056203 (2007).
%
\\ 
%
C.~Manchein, M.~W.~Beims, and J.~M.~Rost, ``Characterizing the dynamics of higher dimensional
nonintegrable conservative systems'', {\em Chaos}, {\bf 22}, 033137 (2012).
%
\\ 
%
S.~Sawada and T.~Taniguchi, 
``Chaos and ergodicity of two hard disks within a circular billiard'', 
{\em Phys. Rev. E}, {\bf 88}, 022907 (2013). 
%
\\ 
%
K.~Kanno and A.~Uchida, ``Finite-time Lyapunov exponents in time-delayed nonlinear dynamical
systems'',  {\em Phys. Rev. E}, {\bf 89}, 032918 (2014). 
%
\\ 
%
D,~Paz\'o, J.~M.~L\'opez, and A.~Politi, 
``Diverging Fluctuations of the Lyapunov Exponents'' 
{\em Phys. Rev. Lett.},  {\bf 117}, 034101 (2016). 
%
\\ 
%
M.~Hanada, H.~Shimada, and M.~Tezuka, ``Universality in Chaos: Lyapunov Spectrum and
Random Matrix Theory'', arXiv:1702.06935.
